---
title: "Itom 6253 Programming for Analytics Fall A 2021 - Assignment 1 - Tanaka Tembo_49699358"
output:
  pdf_document: default
  html_document: default
  date: "Fall A 2024"
---
<!-- name of chunk = set_global_options and include chunk in doc -->
```{r set_global_options, echo=TRUE}
 knitr::opts_chunk$set(echo = TRUE)
```
The raw data file for this assignment is full of problems. There are missing values, outliers and invalid values (e.g., negative revenue). There are extra commas at the end of each line that cause the creation of empty nonsense variables and a few lines are missing a comma to separate the last two variables. Many lines also have zero purchases for all products.

Before doing any analytics, we must first scrub the raw data to solve these problems. We make use of the **dplyr** and **tidyr** packages for many of the scrubbing tasks.

### Instructions
This template requires that you replace all of the instances of "***" with your own code. Sometimes you will need to supply just a single keyword or argument to a function, sometimes a single line of code, and sometimes you need to supply multiple lines of code. Do your best and try to run the entire markdown document without errors for submission.

Once you have finished, knit the document to a pdf file and submit your version of the markdown file (*.rmd) and the associated pdf to the assignment on Canvas.


```{r echo=TRUE}

library(tidyr)
library(dplyr)
library(ggplot2)
library(readxl)

# Importing cleaning of the given data set
cbc <- as.data.frame(
  read_excel("C:\\Users\\sh0556\\OneDrive - Generation Health\\Desktop\\Charles_BookClub_modified.xls")
)

cbc <- cbc[, 1:18]
data_structure <- str(cbc)
data_summary <- summary(cbc)

# This is the printing of the data and its summaries
print(data_structure)
print(data_summary)
```

#### Count the unique values of each variable.
```{r}
# This code is for counting of distinct values in each of the given columns on the data set
cbc_counts <- cbc %>%
  summarise_all(~ n_distinct(.))

# Printing the counts of distinct values
print(cbc_counts)
```

#### Convert Gender and Florence to Factors

```{r}
cbc$gender <- factor(cbc$Gender, levels = c(0, 1), 
                     labels = c("Female", "Male"))

cbc$florence <- factor(cbc$Florence, levels = c(0, 1), 
                       labels = c("No", "Yes"))
```


#### Identify the numeric columns for which outlier detection is desired

```{r}
# I will start by checking if the needed columns do exist in the data set as a confirmation
if (!("M" %in% names(cbc)) || !("R" %in% names(cbc))) {
  stop("Columns M and R do not exist in the data frame.")
}

# Checking the type of variables M and R 
is_numeric_M <- is.numeric(cbc$M)
is_numeric_R <- is.numeric(cbc$R)
            cat("Is column M numeric? ", is_numeric_M, "\n")
            cat("Is column R numeric? ", is_numeric_R, "\n")

```

#### Find outliers and set them to missing
Note the use of the *anonymous* function in the following code:
```{r}
# Convert column M to numeric and set non-numeric or negative values to NA
cbc$M <- as.numeric(cbc$M)
cbc$M <- ifelse(cbc$M <= 0, NA, cbc$M)  

# Convert column R to numeric, setting non-numeric values to NA
cbc$R <- as.numeric(cbc$R) 
cbc$R <- ifelse(cbc$R <= 0, NA, cbc$R)  
```


#### Summary also counts the number of missing values
```{r}
summary(cbc)
```

#### Identify variables for which imputation of missing values is desired
```{r}
# Check for missing values in the entire data frame 
missing_summary <- sapply(cbc, function(x) sum(is.na(x)))

# Identify columns with missing values
columns_with_na <- names(missing_summary[missing_summary > 0])

# Print the columns that require imputation
cat("Columns with missing values that require imputation:\n")
print(columns_with_na)
```
#### Impute missing values of columns with missing values

Here's another *anonymous* function use:
```{r}

cbc$M[is.na(cbc$M)] <- mean(cbc$M, na.rm = TRUE) # Mean imputation for M
cbc$R[is.na(cbc$R)] <- mean(cbc$R, na.rm = TRUE) # Mean imputation for R

# Checking NA values that will possibly remain 

cat("Number of NA values in M after imputation:", sum(is.na(cbc$M)), "\n")
cat("Number of NA values in R after imputation:", sum(is.na(cbc$R)), "\n")

summary(cbc) 
```

#### Delete rows for which there are no books purchased.
```{r}
cbc_no_zeroes <- cbc[rowSums(cbc[, c("ChildBks", "YouthBks", "CookBks", "DoltYBks", "RefBks", "ArtBks", "GeogBks")] > 0) > 0, ]
summary(cbc_no_zeroes)
nrow(cbc)
nrow(cbc_no_zeroes)
```

#### Sum the purchases of each book type.
```{r}
cbc_sums <- cbc %>% summarise(across(c(ChildBks, YouthBks, CookBks, DoltYBks, RefBks, ArtBks, GeogBks, ItalCook, ItalHAtlas, ItalArt), sum))
cbc_sums
```

### Histogram plot of numeric variables

```{r}
library(psych)
multi.hist(cbc[, c("M", "F", "R", "FirstPurch")], nrow=2, ncol=2, global = FALSE)
```

### Bar plot of book type sums

```{r}
# This is pivoting my data set
cbc_pivot <- pivot_longer(cbc_sums, cols = everything())  
names(cbc_pivot) <- c("Type", "Sum")

# Creating of a bar plot code
ggplot(data = cbc_pivot, aes(x = Type, y = Sum)) +
  geom_bar(stat = 'identity') +
  labs(title = "Sum of Book Types", x = "Book Type", y = "Total Purchases") +
  theme_minimal()
```

### A custom function for calculating 4 moments

```{r}
calcfourstats <- function(x) {
  mu <- round(mean(x, na.rm = TRUE), 2)
  sigma <- round(sd(x, na.rm = TRUE), 2)
  skew <- round(skewness(x, na.rm = TRUE), 3)
  kurt <- round(kurtosis(x, na.rm = TRUE), 2)
  result <- data.frame(mu, sigma, skew, kurt)
  return(result)
}

# This shows the list of column names and corresponding variable names
columnNames <- c("M", "R", "F", "FirstPurch")
varList <- c("Monetary Value", "Last Purchase", "Frequency", "First Purchase")
results <- data.frame()

# Calculating the moments for each column
for (col in columnNames) {
  results <- rbind(results, calcfourstats(cbc_no_zeroes[[col]]))
}

# This is the setting of Set row names for the results
rownames(results) <- varList

# Print results
print(results)
```

### Creating RFM factors
#### Calculate HML cutoffs for RFM

```{r} 

library(dplyr)

# Calculate HML cutoffs for R, F, and M
cbc_rfm_cutoffs <- data.frame(lapply(cbc_no_zeroes[c("R", "F", "M")], 
  function(x) {
    quantile(x, probs = c(0, 0.33, 0.67, 1), na.rm = TRUE)
  }))

# Extract cutoffs for R, F, and M
cutoffs <- as.list(cbc_rfm_cutoffs)

# Create RFM categories based on the calculated cutoffs
cbc_rfm_categories <- cbc_no_zeroes %>%
  mutate(
    r_category = case_when(
      R <= cutoffs$R[2] ~ "Low",    # 0 to 33%
      R <= cutoffs$R[3] ~ "Medium", # 34% to 66%
      TRUE ~ "High"                 # 67% to 100%
    ),
    f_category = case_when(
      F <= cutoffs$F[2] ~ "Low",
      F <= cutoffs$F[3] ~ "Medium",
      TRUE ~ "High"
    ),
    m_category = case_when(
      M <= cutoffs$M[2] ~ "Low",
      M <= cutoffs$M[3] ~ "Medium",
      TRUE ~ "High"
    )
  )

# Create 3x3 cross tabulation tables of means for M by frequency and recency levels
mean_tables <- list()

for (m_level in c("Low", "Medium", "High")) {
  mean_table <- cbc_rfm_categories %>%
    filter(m_category == m_level) %>%
    group_by(f_category, r_category) %>%
    summarise(mean_m = mean(M, na.rm = TRUE)) %>%
    ungroup() %>%
    pivot_wider(names_from = r_category, values_from = mean_m, values_fill = list(mean_m = NA))

  mean_tables[[m_level]] <- mean_table
}

# Print the mean tables for each level of M
for (level in names(mean_tables)) {
  cat("Mean table for M level:", level, "\n")
  print(mean_tables[[level]])
  cat("\n")
}
```

## Verify results and test subsetting    
```{r}
# Accessing of the 33rd percentile of M using index 
percentile_33_M <- cbc_rfm[2, "M"]  
cat("The 33rd percentile of M is:", percentile_33_M, "\n")
```

# Create three new variables for HML quantiles of RFM variables
```{r}
library(dplyr)

cbcRFM <- cbc_no_zeroes %>%
  mutate(
    rRFM = if_else(R <= cbc_rfm["Low Quantile", "R"], "L",
                   if_else(R >= cbc_rfm["High Quantile", "R"], "H", "M")),
    
    fRFM = if_else(F <= cbc_rfm["Low Quantile", "F"], "L",
                   if_else(F >= cbc_rfm["High Quantile", "F"], "H", "M")),
    
    mRFM = if_else(M <= cbc_rfm["Low Quantile", "M"], "L",
                   if_else(M >= cbc_rfm["High Quantile", "M"], "H", "M"))
  )

# View the updated data set with RFM factors
head(cbcRFM)
```

#Convert the new HML variables into ordered factors
```{r}
library(dplyr)

cbcRFM[c("rRFM", "fRFM", "mRFM")] <- data.frame(lapply(cbcRFM[c("rRFM", "fRFM", "mRFM")], 
  function(x) {
    factor(x, levels = c("L", "M", "H"), ordered = TRUE)
  }))

# Option of Viewing the updated data-set
head(cbcRFM)
str(cbcRFM)

# Summarizing the dataset as a whole
sumTable <- cbcRFM %>% 
  group_by(rRFM, fRFM, mRFM) %>%
  summarize(meanM = round(mean(M, na.rm = TRUE), 2), .groups = 'drop')
  print(sumTable)
```

#### Make three tables, one for each level of factor mRFM

```{r, echo=TRUE, message=FALSE, warning=FALSE}
for (i in c("L", "M", "H")) {
  shortTable <- xtabs(meanM ~ rRFM + fRFM, sumTable %>% filter(mRFM == i)) 
  print(paste('Monetary Value Segment =', i))
  print(shortTable)
  cat("\n") 
}
```

### Median monetary value per visit by gender

```{r}
# Calculate the median monetary value per visit by gender
visitValue <- cbcRFM %>%
  group_by(Gender) %>%
  summarize(medianM = round(median(M / F, na.rm = TRUE), 2))

# Print the results
visitValue
```

#### Bubble plot of M by R by Gender

```{r}
library(ggplot2)

# Bubble plot of M by R, colored by gender
ggplot(cbc, aes(x = R, y = M, 
                 col = factor(gender, labels = c("Female", "Male")), 
                 size = FirstPurch)) +
  geom_point(alpha = 0.2) +
  labs(x = "Recency", y = "Monetary Value") +
  facet_wrap(~ factor(gender, labels = c("Female", "Male")), 
             labeller = label_parsed) +
  theme(legend.position = "bottom", 
        legend.box = "vertical", 
        legend.key = element_rect(colour = "white", fill = "white")) +
  theme(strip.text.x = element_blank())  # Remove facet label for gender
```

